---
title: 'Reading Files: Beyond the Basics'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(fs)
library(janitor)
```

Reading files into R is often the start of a data analysis, and there are a number of tools to help make data import as efficient as possible.

## Bread and butter data import with the *readr* package

Arguably the best "out of the box" package for data import from formatted plain text files is [readr](http://readr.tidyverse.org/articles/readr.html), which is one of the packages in the tidyverse. The syntax for function names in this packages is very straightforward: `read_csv()` indicates a read operation on a csv file type. Tab-delimited files can be read in with `read_tsv()`. The most generic file reading function in this package is `read_delim()`, which allows you to indicate the delimiter in the file to separate columns. 

A common challenge in importing data is ensuring that the data type for a given column aligns to how you expect to work with the data. The functions in the readr package will scan the first 1000 entries by default and guess the column type based on those entries. This generally helps decrease the amount of effort required to read in data since you don't have to explicitly specify data types for each column. However this behavior does not guarantee the intended outcome for a specific field in your data set. For example, if you are importing a field that you expect to have numerical values but there are some entries with text values in the first 1000 rows, the data type for that field will be set to a character. To help navigate this issue, readr functions also provide a syntax for explicitly defining column types:

```{r, eval = FALSE}
# purely a dummy example, not executable!
imaginary_data_frame <- read_csv(
  "imaginary_file.csv",
  col_types = cols(
    x = col_integer(),
    y = col_character(),
    z = col_datetime()
  )
)
```

In addition to the data types in the example, there are a number of other formats supported by the `col_` syntax: logical, double, factor (need to specify levels), date, time, datetime. Another advantage of these functions: on import you will see that they actually explicitly tell you how the columns were parsed when you import (as we'll see in the exercise).

**Exercise 1**

Now let's run through using the readr function for a csv:
1. Use the `read_csv()` function to read the "2017-01-06_s.csv" file within your "data" folder (place that folder into your "exercises" folder with this file if you haven't already) into a data frame. The file is within the "data" folder so you will need to provide a path to that files that includes the folder.

```{r}
readr_load <- 
```

2. What is the internal structure of the object? (Hint: use the `str()` function.)

```{r}

```

3. Summarize the data.

```{r}

```

4. Finally, let's follow some best practices and explicitly define columns with the `col_types` argument. We want to explicitly define compound_name and sample_type as factors. Note that the `col_factor()` expects a definition of the factor levels but you can get around this by supplying a `NULL`. Then run a summary to review the data.

```{r col_types}
readr_load_factors <- read_csv( ,
  col_types = cols(
    
    
    )
  )
summary( )
```

**End Exercise**

## Dealing with Excel files (gracefully)

You may have broken up with Excel, but unfortunately many of your colleagues have not. You may be using a little Excel on the side. (Don't worry, we don't judge!) So Excel files will continue to be a part of your life. The [readxl package](http://readxl.tidyverse.org/) makes it easy to read in data from these files and also offers additional useful functionality. As with the other file reading functions, the syntax is pretty straightforward: `read_excel("file_name.xlsx")`. Different portions of the spreadsheet can be read using the `range` arugment. For example a subset of rows and columns can be selected via cell coordinates: `read_excel("file_name.xlsx", range = "B1:D6")` or `read_excel("file_name.xlsx, range = cell_cols("A:F")`.

Excel files have an added layer of complexity in that one file may have multiple worksheets, so the `sheet = "worksheet_name"` argument can be added to specify the desired worksheet: `read_excel("file_name.xlsx", sheet = "worksheet_name")`. If the file has multiple worksheets and you don't specify, it will read the first one by default. In case you haven't opened an Excel file manually, there is also a helpful function to list the sheets in a file: `excel_sheets()` takes the path of the file as the argument and returns the list of sheets.

**Exercise 2**

You might be able to guess what comes next: we'll examine and read in an Excel file.
1. Use the `excel_sheets()` function to examine the "2017-01-06.xlsx" file within the "data"
2. Use the `read_excel()` function to read this file into a data frame
3. View a summary of the imported data
4. Now read in only the first 5 columns of the samples worksheet using the `worksheet` and `range` parameters
5. Review the first 6 lines of the imported data

```{r readxl}
# complete the function calls and code below
excel_sheets() # expecting a file path as input

readxl_load <- read_excel( ) 
## insert additional code here

readxl_load_subset <- read_excel( , sheet = , range = )
## insert additional code here
```

**End Exercise**

If you are dealing with Excel data that is not a traditional tabular format, the [tidyxl package](https://cran.r-project.org/web/packages/tidyxl/vignettes/tidyxl.html) is useful to be aware of. We will not cover it in this course but it is worth reading up on if you ever have to analyze a pivot table or some other product of an Excel analysis.


## Importing dirty data

Very often the first set of operations you may want to perform on a data set that's imported is data cleaning. One package that can be very helpful for straightforward data cleaning activities is cleverly and appropriately named [janitor](https://github.com/sfirke/janitor). The quick take home in terms of useful functions from this package:
- `clean_names()` will reformat column names to conform to the tidyverse style guide: spaces are replaced with underscores & uppercase letters are converted to lowercase
- `tabyl()` will tabulate into a data frame based on 1-3 variables supplied to it
- `get_dupes()` returns duplicate records given a set of one or more variables
- empty rows and/or columns are removed with `remove_empty()`

Let's take these functions for a spin using our data set. First let's review the first few lines of data after cleaning the column names:
```{r janitor}

# readxl_load <- read_excel("data/2017-01-06.xlsx")

head(readxl_load)
readxl_load_cleaned <- readxl_load %>%
  clean_names()
head(readxl_load_cleaned)
```



## Iteration: importing multiple files at once

One of the most compelling reasons to learn how to program is being able to expand your ability to automate or effortless repeat common actions and workflows. In most research and clinic lab environments, the data that people deal with day-to-day is not neatly stored in an easy-to-use database. It is often spread out over a series of messy spreadsheets that might be associated with one batch of data, one day of data, one week of data, or some variant. While the best practice for that scenario is probably to build a database to store the data, that requires a good amount of overhead and some expertise. By taking advantage of iteration in R, you can dump similiarly formatted files into data frames/tibbles.

The [purrr package](https://purrr.tidyverse.org) has a variety of `map()` functions that are well-explained in the [iteration chapter](http://r4ds.had.co.nz/iteration.html) of R for Data Science. The `map()` functions take a vector as an input, applies a function to elements of the vector, and returns a vector of identical length to the input vector. There are a number of map functions that correspond to the data type of the output. For example, `map()` returns a list, `map_int()` returns a vector of integers, `map_chr()` returns a character vector, and `map_dfr()` returns a data frame. These are very similar to the `apply()` family of functions but there are some advantages of the purrr functions, including consistent compabibility with pipes and more predictable output data types. 

How does this work? Let's take a simple example right out of the R for Data Science text. We'll start with a tibble (tidyverse version of data frame) consisting of 4 variables (a through d) with 10 observations from a normal distribution.

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
df
```

We want to treat each variable as a vector and perform a calculation on each. If we want to take the mean of each and want the output to have a double data type, we use `map_dbl()`:

```{r}
df %>%
  map_dbl(mean)
```

That is a pretty simple example but it captures the types of operations you can you do by iterating through a data set. For those of you who are familiar with for loops, the map functions can offer similar functionality but are much shorter to write and straight-forward to understand.

Earlier in this lesson we discussed file reading functions, with the recognition that many data analysis tasks rely on flat files for source data. In a laboratory running batched testing such as a mass spectrometry lab, files are often tied to batches and/or dates and named correspondingly. If you want to analyze a set of data over multiple batches, you may find yourself importing data from each individually and stitching together the data using a function like `bind_rows()` (we will discuss this function in a future lesson). The `map()` functions (often `map_dfr()` specifically) can automate this process and save you a lot of time. There are a few prerequisites for this to work, though:

- the underlying file structure must be the same: for spreadsheet-like data, columns must be in the same positions in each with consistent data types
- the files must have the same file extension
- if there are multiple different file types (with different data structures) mixed in one directory, the files must organized and named in a way to associate like data sets with like

Let's say we have a large mass spec data set consisting of a series of daily data files that are grouped into batches, samples, and peaks data, with suffixes of "_b", "_s", and "_p", respectively. Let's read all of the sample data into one data frame (technically a tibble). We are going to use the `read_csv()` function since the files are csvs. To use the `map_dfr()` function, we need to supply a vector as input - in this case, a vector of file names. How do generate that input vector?

This requires us to use functions designed for file system operations such as listing the files in a directory. There are base R functions for this, but we prefer to use the `fs` package, which improves upon some of the idiosyncrasies of base R for manipulating paths, files, and directories. This is also a tidyverse compliant package, so it works with pipes and the `map()` functions described above.

- First, we use `dir_ls()`, which produces a character vector of file names for a directory, which is specified in the first argument (path = ) of the function. The function allows a pattern argument (glob = ) which you can supply with a text string for it to match against - all of the sample files end in "_s.csv".

- Next, we feed that character vector to `map_df()`, which takes the `read_csv()` function as its argument. We are specifying that want each file in the "data" directory read in using read_csv and combined into a single dataframe to create the all_samples object. With the map family of functions, there is no need to include the parentheses in the function name if there aren't arguments.

```{r samples}
sample_file <- dir_ls("data", glob = "*_s.csv") %>%
  map_dfr(read_csv) %>%
  clean_names()

# always check your output!!
str(sample_file)
summary(sample_file)
```

 

**Exercise 3**

Now it is your turn to apply this same code for the batches and peaks file types to generate dataframes for those data.
1. Use the `dir_ls()` function to read the filenames for the batches and the peaks types of files into vectors
2. Read each type of these files into a single dataframe using `read_csv()` and `map_dfr()`
3. Clean the column names
4. Review the structure of the new dataframes

```{r batches}
batch_file <- dir_ls("data", glob = "") %>%
  map_dfr() %>%
## insert additional code here

## insert additional code here
```

```{r peaks}
peak_file <- dir_ls("data", glob = "") %>%
  map_dfr() %>%
## insert additional code here

## insert additional code here
```

**End Exercise**


We do this separately for each file type since the structures are different across the samples, peaks, and batches files. The end goal is for those dataframes to be written to the matching tables in our SQLite database.

If you weren't already aware of this solution or another for reading in multiple files at once, the purrr package is an extremely handy tool for doing this. Just be aware of the requirements for doing this, and **always check the output**. You do not want to automate a bad or broken process!

## Summary

- readr functions such as `read_csv()` are faster than base R functions and do not automatically convert strings to factors
- The readxl function `read_excel()` reads Excel files and offers functionality in specifying worksheets or subsets of the spreadsheet
- The janitor package can help with cleaning up irregularly structured input files
- The purrr package has useful tools for iterating that can be very powerful when coupled with file reading functions